Nengo: A Python tool for building large-scale functional brain models

### 2013-09-27 nengo.tex
Major rewrites

\def\journal{Neuroinformatics}
\def\DOI{}
\def\articleType{Methods Article}
\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Bekolay {et~al.}}
\def\Authors{Trevor Bekolay\,$^{1,*}$, James Bergstra\,$^{1}$,
  Xuan Choo\,$^{1}$, Travis DeWolf\,$^{1}$, Eric Hunsberger\,$^{1}$,
  Daniel Rasmussen\,$^{1}$, Terrence C. Stewart\,$^{1}$
  and Chris Eliasmith\,$^{1}$}
\def\Address{$^{1}$Computational Neuroscience Research Group,
  Center for Theoretical Neuroscience,
  University of Waterloo, Waterloo, ON, Canada}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name
% and city zip code) and email of the corresponding author
\def\corrAuthor{Trevor Bekolay}
\def\corrAddress{David R. Cheriton School of Computer Science,
  University of Waterloo, 200 University Avenue West,
  Waterloo, ON, N2L 3G1, Canada}
\def\corrEmail{tbekolay@uwaterloo.ca}

@@

It is often said that neuroscience
is data-rich but theory-poor.
Although computational neuroscience
has grown significantly,
this phrase continues to be repeated
due to a lack of satisfying
theories of the brain.
Much work has been put into simulators
that attempt to recreate neural
data in precise detail,
but behavior has not emerged
from data-driven large scale models.
Similarly, theoretical frameworks
that involve a generic algorithm
replicated many times
(e.g., the memory-prediction framework \cite{TODO}
and the pattern recognition theory of mind \cite{TODO})
have not been shown to solve problems
outside of those that
the generic algorithm alone can solve.

Nengo is a neural simulator
based on a theoretical framework proposed
by Eliasmith \& Anderson \citeyearpar{TODO}
called the Neural Engineering Framework
(NEF).
Nengo was recently used to implement Spaun,
currently the world's largest functional brain model
\cite{TODO}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight high-level cognitive tasks
including list memory and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed
and the details of that task
(e.g., $A3[1234]$ instructs Spaun
to remember the list $1234$).
It then provides its answer by
sending motor commands to a simulated arm model,
tracing out digits
(e.g., $1234$ if asked to recall the remembered list).

While the tasks Spaun performs are diverse,
all tasks use a common set of
functional modules that we map
onto the brain areas hypothesized
to perform those functions.
Like data-driven large scale models,
Spaun is able to match experimental data
across the neuroscientific spectrum,
from the single cell to the behavioral level.
However, instead of constructing the model
based on data and hoping behavior emerges,
we construct the model
based on behavior
and see that the data emerges.
This is made possible by a theoretical framework
like the memory-prediction framework \cite{TODO},
but unlike it and other frameworks,
the NEF enables Spaun to employ
many different algorithms
based on the behavior being performed,
rather than defining a single algorithm
and hoping that replicating it
many times will enable these behaviors.

In implementing Spaun,
we have provided strong support
that the NEF is beginning to fill the theory void in neuroscience.
However, we have also pushed Nengo
to the limits of its ability to construct and simulate
large-scale neural models.
In order to continue to evaluate the NEF
as a theory of the brain
by building larger and more complicated models,
we have begun the next generation
of Nengo tools with a focus on simplicity,
extensibility, and speed.

@@

\begin{table}[!t]
\processtable{Nengo tools\label{tab:nengo-vers}}
{\begin{tabular}{p{2.6cm} p{2.7cm} l p{7cm}} \toprule
\textbf{Name} & \textbf{Implementation language}
  & \textbf{Code repository} & \textbf{Purpose} \\\midrule
JavaNengo simulator & Java & \texttt{nengo\_java}
  & Simulates many types of spiking neuron models.
  Includes methods to construct and simulate models using the NEF. \\
JavaNengo GUI & Java & \texttt{nengo\_java\_gui}
  & Graphical interface to JavaNengo simulator.
  Allows modellers to drag-and-drop components built with the NEF. \\
JavaNengo scripting interface & Jython & \texttt{nengo\_jython}
  & Scripting interface to JavaNengo simulator.
  Provides shortcuts to classes and functions in the simulator.
  Also accessible through and interacts with the JavaNengo GUI. \\
PyNengo & Python & \texttt{nengo}
  & Constructs and simulates models using the NEF. \\
PyNengo OpenCL simulator & Python+PyOpenCL
  & \texttt{nengo\_ocl}
  & Simulates models constructed with PyNengo. \\\botrule
\end{tabular}}{}
\end{table}

JavaNengo is a suite of tools providing
a neural simulator and graphical user interface (GUI)
implemented in Java,
and a scripting interface implemented in Jython
(see Table~\ref{tab:nengo-vers} for more detail).
JavaNengo's simulator was created
with the intention of being a general
neural simulator that included
the methods of the NEF
as one of many different
methods of creating neural models.
As a result, its implementation
of important NEF components
are nested in deep class hierarchies,
adding unnecessary layers of complexity
for developers wishing
to extend those NEF components.
The addition of synaptic plasticity rules,
for example, involved adding or significantly modifying
25 Java classes \cite{TODO}.\footnote{Changelog at
  \url{https://github.com/ctn-waterloo/nengo_1.4/commit/93eee51147545f7ce30ae9919809e3072dc15728}.}

@@

The JavaNengo scripting interface,
described in \cite{TODO},
provided a simpler interface
to the NEF-specific portions of the simulator.
This greatly enhanced productivity,
and made it possible
to create many models with very little code.
Modellers quickly started using
the scripting interface
as their primary tool,
but still needed to access the GUI
for interactive model inspection,
and still needed to modify the simulator
to add new functionality
(e.g., new neuron models).
This is because Jython makes it easy
to inspect and create Java objects,
but does not make it easy to
write Java code to
inspect and create Jython objects.
Modellers, therefore, had to be
proficient in Java whether
they used the scripting interface or not.
As the set of neural modellers
that are proficient with both Java and Python
syntax is small,
the scripting interface still lacks
convenient ways to record the results of simulations,
and to define complex experimental setups.

Finally, perhaps the most significant limitation
of the JavaNengo tools is its performance.
While Java has good support for threads,
it does not have well-supported libraries
for doing mathematical operations on large matrices quickly.
Additionally, it is difficult to interface with
platform-specific tools such as
graphical processing units (GPUs),
which can perform the types of computations
that these neural models require quickly and in parallel.

Addressing all of these issues
in the Java codebase would involve
significant enough changes
that we decided to start from scratch instead.
Since the scripting interface
already used Python syntax,
and we were using Python
and its scientific libraries
for analyzing simulation results,
we targeted the CPython interpreter
as the platform for the new generation of Nengo tools.

@@

PyNengo, which we describe in more detail
in the remainder of this paper,
is a Python package for
defining and simulating
neural models using the NEF.
It targets the CPython interpreter,
meaning that it integrates seamlessly
with other scientific Python tools.
Nengo models can used
in arbitrary CPython programs,
opening up possibilities
for using neurally inspired algorithms
in applications and games.

PyNengo's object model has been radically simplified,
making it easy to
document, test, and modify those objects.
It also includes a scripting interface to those objects
based on JavaNengo's scripting interface \cite{TODO},
but with several modifications
to allow for easier incorporation
of new neuron models and learning rules.
Additionally, PyNengo's scripting interface
contains methods to record simulated data
and access that data after simulation.

Model creation and simulation
have been completely decoupled,
allowing for models to be exported
to other simulators,
such as those accessible with PyNN.
Simulators can also be written
to leverage specific platforms.
Currently, we have implemented
a platform-independent simulator
that uses NumPy for vectorized computations,
and a PyOpenCL-based
simulator that takes full advantage of GPUs
and multicore CPUs.

PyNengo does not currently
replace all of the JavaNengo tools.
Most notably, there is no graphical interface
to PyNengo, and therefore little ability to interact
with simulations as they are running.
However, we are confident that these tools
can be created, and will be much easier
to implement compared to JavaNengo's GUI.

We believe that PyNengo overcomes
all of the significant issues
with JavaNengo.
It can simulate the majority of models
currently simulated by JavaNengo,
using 11\% as many lines code,
and depending on
only three external libraries instead of ten.
The OpenCL simulator can simulate
a large scale network TODO times
faster than JavaNengo's simulator.
This, along with a codebase
that is easier to understand and extend,
provides a platform
that can simulate larger and more complex
models than Spaun,
and can therefore further
test the NEF as a theory of the brain.

@@

TODO; make sure it meshes with 3.1's operators

\begin{equation} \label{TODO}
  TODO
\end{equation}

@@

\section{Model definition}

One of the strengths of the NEF
is that it allows neural modellers to think
at a higher level
than is normally associated with neural models.
With the NEF, a modeller decides
what information is being represented
by populations of neurons and how that information
is being transformed through the connections
between populations;
this contrasts with modelling decisions
that are typically thought to be important,
such as the morphology of each neuron,
and the distribution of connection weights
between populations.
Instead, these decisions are made---often
automatically---as a direct result
of the information
being represented and the transformations
being performed on that information.
This situates Nengo as a
``neural compiler'' that can
take a functional model description
and produce a low-level neural model
that can be simulated
with any neural simulation software.

To emphasize Nengo's role as a neural compiler,
Nengo allows models to be defined
in a simulator-agnostic manner,
allowing for functionally identical models to be run
on different software and hardware platforms.

\subsection{Nengo objects} \label{sec:objects}

A Nengo model is a collection
of Nengo objects connected together,
and probed in order to collect data
during simulation.
The \texttt{Model} object is primarily a container
for Nengo objects,
but also provides methods
that simplify creating, connecting,
and probing these objects.
Basic Nengo models
can be created solely through interacting
with the \texttt{Model} object.
For more advanced models,
these objects can be instantiated
directly and then added to the model.

There are two types of objects
defined in the main scripting interface.

\paragraph{Ensembles}
The most common is the \texttt{Ensemble},
which is a population of neurons
that represents information
in the form of a real-valued vector.
As described in \eqref{TODO},
this can be done
with any neuron model, $G$.
An ensemble must be given as arguments
its dimensionality
(i.e., the length of the vector it represents),
and an object that describes
the population of neurons.
For example, \texttt{LIF(50, tau\_rc=0.02)}
describes a population
of 50 leaky integrate-and-fire neurons
with an RC time constant of 0.02 seconds.
Neurons are defined in this symbolic manner
so that each simulator can compute
this population's nonlinear function
however it sees fit.

Other attributes of the \texttt{Ensemble},
such as its encoding weights, $\mathbf{e_i}$,
the maximum firing rate of the neurons,
and so on, can also be specified.
This is usually done according
to what is known about
the neurobiology of the area being modelled;
some areas of the brain regularly spike
at rates above 200Hz,
while others rarely spike above 10Hz.
If these attributes are not set,
then they will be randomly selected
from distributions that describe
what is known about pyramidal cells in neocortex.

\paragraph{Nodes}
A \texttt{Node} tracks any information
that is not represented by an ensemble.
In the most common case,
nodes provide input signals
that drive ensembles of sensory neurons.
In a more general sense,
nodes represent the experimental environment
in which a neural model exists.
More practically,
nodes can be used to directly compute
parts of a neural model
for debugging purposes.

@@

Nodes allows many of the special-purpose
routines that are necessary in many real-world
models to be integrated
as core components of a Nengo model.

\paragraph{Connections}

@@

The most important type of connection
in Nengo is the \texttt{DecodedConnection}.
This connection implements
the transformation principle equations,
\eqref{TODO}.
In other words, the \texttt{DecodedConnection}
allows ensembles to project
the information they're representing---or
a transformation of that information---to
another ensemble or node.
This functionality is what enables Nengo models
to appear conceptual,
even though the underlying implementation
still connects populations of neurons
together with connection weights.
%TODO: reword? is it important that it's a factored weight matrix

\paragraph{Probes}

Ideally, the entire state of a simulation
would be recorded on each timestep,
and stored with no overhead to the simulator.
In practice, recording data
is a costly part of most simulations,
and often what is simulated must also change
if the signals being probed must be filtered.
For this reason, probes are also
a core component of a Nengo model.

In general, Nengo objects and connections
each have several signals that can be probed.
For example, an ensemble's decoded value,
or the underlying neurons' membrane voltages
can be probed, among other signals.
The underlying logic for the probe
leverages connection logic,
making it easy to probe any part of an object.

Since probes are also described
at a symbolic level,
the underlying implementation
can output probed data in many different formats.
Currently, simulators store probed data
directly in memory,
but adding the ability to store data
in files or to stream probed data
directly to sockets will be added.

\section{Model simulation} \label{sec:simulators}

Once a model has been fully defined,
a simulator can be created to
run that model.
Simulators can be created
that work directly with the
objects described in section \ref{sec:objects}.
However, in the simulators that we have implemented
as reference, the model goes through a
two-stage build process in order to
simulate models as fast as possible.
In the first stage,
the model is ``built;''
i.e., the objects in the model are mapped to a simpler
set of objects that represent the computations
that the simulator must perform.
Those computations are analyzed in order to
schedule them efficiently.
In the second stage,
the simulator performs the computations
that have been scheduled on each timestep.

This two-stage process has enabled us
to implement a more efficient simulator
that is only aware of
the objects that are created
during the build process.
In the future,
new simulators can target either
the high level Nengo objects,
or the low level simulator objects
depending on the constraints of the platform.

\subsection{Reference build process}

\begin{figure}
\begin{center}
  \includegraphics[width=0.6\textwidth]{sim}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:sim} Figure \arabic{figure}.}{
   Sim}
\end{figure}

The first stage of the reference implementation
maps all of the Nengo objects
down to a set of signals and operators.

\paragraph{Signals}
A \texttt{Signal} represents any number that
will be used by the simulator.
Each high level Nengo object contains
several signals;
for example, an ensemble contains signals
that represent the high-level input
signal that will be encoded
to input currents (see equation \eqref{TODO}),
and the encoding weights
($\mathbf{e_i}$; see equation \eqref{TODO}).
It also contains a neural population,
which contains signals that represent
input currents
($J(t)$; see equation \eqref{TODO}),
bias currents ($J^{bias}$; \eqref{TODO}),
membrane voltages ($V$; \eqref{TODO}),
and refractory times for each cell.
These signal are symbolic,
and contain only information
about the size of the signal that will be represented,
and optionally an initial value.
Simulators may therefore represent
these signals however they want internally.

In some cases, determining
the initial values of some signals
may involve some computation.
In particular, solving for decoding weights
(equation \eqref{TODO}) requires
solving a least-squares minimization problem.
This computation is done in this stage
of the build process,
as it is computationally intensive,
and may be implemented
differently by some simulators.

An important feature of signals
is that parts of a signal
can be accessed with Python slice syntax,
which results in a view of the original signal.
For example, when connecting to
a subset of a population of neurons,
current would be injected
in a view of the overall input current signal
for that population,
rather than creating a new signal,
or partitioning the population's signal
into smaller chunks.
Grouping related signals together
and exposing them through views
also provides hints to the simulator
that can be used to organize signals
efficiently in memory.

\paragraph{Operators}
Operators represent computations
to be performed on signals on each timestep.
Once the model has been built,
only a small set of mathematical
operations are necessary for simulation.
\begin{enumerate}
  \item \textbf{Reset} assigns a constant value to a signal.
    This is typically used to set a signal to 0
    at the start of a timestep.
  \item \textbf{Copy} assigns the value
    of one signal to another signal.
    This is typically used instead of \textbf{Reset}
    in situations where the signal's value
    at the start of a timestep should
    be the value of another signal.
  \item \textbf{DotInc} increments a signal by
    the dot product of two signals;
    i.e., $Y \leftarrow Y + AB$.
    This is a common operation
    done in many situations,
    such as decoding a signal (see equation \eqref{TODO}).
  \item \textbf{ProdUpdate} scales a signal
    and adds the dot product of two signals to it;
    i.e., $Y \leftarrow BY + AX$.
    This is done when temporally filtering a signal.
  \item \textbf{Nonlinearities} include any
    user-defined function computed by a node,
    and the update equation for each neuron model.
\end{enumerate}

It is important to note that all operators
excepts for the nonlinearities are linear,
and therefore can share an implementation
in an underlying simulator,
allowing those operations to happen in parallel.

Operators are created
by Nengo objects during the build process.
For example, an ensemble creates
a \textbf{Reset} operator to reset
its input signal to 0,
and a \textbf{DotInc} operator
to encode that input signal into
input currents injected into
the underlying population of neurons;
i.e., the \textbf{DotInc} computes
$J = \alpha_i \mathbf{e_i} X$ from equation \eqref{TODO}.

@@

The reference simulator starts by
building NumPy arrays
for every signal created during the build process
described above.
It then builds a graph of the
operators created during the build process,
ordering them such that they can be
executed one after another to get the correct answer.

>>>>

\def\journal{Neuroinformatics}
\def\DOI{}
\def\articleType{Methods Article}
\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Bekolay {et~al.}}
\def\Authors{Trevor Bekolay\,$^{1,*}$, James Bergstra\,$^{1}$,
  Xuan Choo\,$^{1}$, Travis DeWolf\,$^{1}$, Eric Hunsberger\,$^{1}$,
  Daniel Rasmussen\,$^{1}$, Terrence C. Stewart\,$^{1}$
  and Chris Eliasmith\,$^{1}$}
\def\Address{$^{1}$Computational Neuroscience Research Group,
  Center for Theoretical Neuroscience,
  University of Waterloo, Waterloo, ON, Canada}
\def\corrAuthor{Trevor Bekolay}
\def\corrAddress{David R. Cheriton School of Computer Science,
  University of Waterloo, 200 University Avenue West,
  Waterloo, ON, N2L 3G1, Canada}
\def\corrEmail{tbekolay@uwaterloo.ca}

@@

It is often said that neuroscience
is data-rich but theory-poor.
Although computational neuroscience
has grown significantly,
this phrase continues to be repeated
due to a lack of satisfying
theories of the brain.
Much work has been put into neural simulators
that attempt to recreate neuroscientific
data in precise detail,
but behavior has not emerged
from data-driven large scale models.
Similarly, theoretical frameworks
consisting of a generic algorithm
replicated many times
(e.g., the memory-prediction framework \cite{TODO}
and the pattern recognition theory of mind \cite{TODO})
have not been shown to solve problems
that the generic algorithm alone cannot solve.

Nengo is a neural simulator
based on a theoretical framework proposed
by Eliasmith \& Anderson \citeyearpar{TODO}
called the Neural Engineering Framework
(NEF).
Nengo was recently used to implement Spaun,
currently the world's largest functional brain model
\cite{TODO}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight high-level cognitive tasks
including list memory and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed
and the details of that task;
for example, sequentially presenting images
containing the characters $A3[1234]$ instructs Spaun
to remember the list $1234$.
It then provides its answer by
sending motor commands to a simulated arm model,
tracing out digits.
If asked to recall the remember list,
it would trace out the digits $1234$.

While the tasks Spaun performs are diverse,
all tasks use a common set of
functional modules that we map
onto the brain areas hypothesized
to perform those functions.
Like data-driven large scale models,
Spaun is able to match experimental data
across the neuroscientific spectrum,
from the single cell to the behavioral level.
However, instead of constructing the model
based on data and hoping behavior emerges,
we construct the model
based on behavior
and see that the data emerges.
This is made possible by a theoretical framework
like the memory-prediction framework \cite{TODO},
but unlike it and other frameworks,
the NEF enables Spaun to employ
many different algorithms
based on the behavior being performed,
rather than defining a single algorithm
and hypothesizing that replicating it
many times will enable these behaviors.

In implementing Spaun,
we provided strong support
that the NEF is beginning to fill the theory void in neuroscience.
However, in doing so, we pushed Nengo
to the limits of its ability to construct and simulate
large-scale neural models.
In order to continue to evaluate the NEF
as a theory of the brain
by building larger and more complicated models,
we have begun the next generation
of Nengo with a focus on speed,
extensibility, and simplicity.

@@

\begin{table}[!t]
\processtable{Nengo tools\label{tab:nengo-vers}}
{\begin{tabular}{p{2.6cm} p{2.7cm} l p{7cm}} \toprule
\textbf{Name} & \textbf{Implementation language}
  & \textbf{Code repository} & \textbf{Purpose} \\\midrule
JavaNengo simulator & Java & \texttt{nengo\_java}
  & Simulates many types of spiking neuron models.
  Includes methods to construct and simulate models using the NEF. \\
JavaNengo GUI & Java & \texttt{nengo\_java\_gui}
  & Graphical interface to JavaNengo simulator.
  Allows modellers to drag-and-drop components built with the NEF. \\
JavaNengo scripting interface & Jython & \texttt{nengo\_jython}
  & Scripting interface to JavaNengo simulator.
  Provides shortcuts to classes and functions in the simulator.
  Also accessible through and interacts with the JavaNengo GUI. \\
PyNengo & Python & \texttt{nengo}
  & Constructs and simulates models using the NEF. \\
PyNengo OpenCL simulator & Python+PyOpenCL
  & \texttt{nengo\_ocl}
  & Simulates models constructed with PyNengo. \\\botrule
\end{tabular}}{Code repositories avilable at \url{http://github.com/ctn-waterloo/}}
\end{table}

JavaNengo is a suite of tools providing
a neural simulator and graphical user interface (GUI)
implemented in Java,
and a scripting interface implemented in Jython
(see Table~\ref{tab:nengo-vers} for more detail).
JavaNengo's simulator was created
with the intention of being a general
neural simulator that included
the methods of the NEF
as one of several
methods for creating neural models.
As a result, its implementation
of important NEF components
are nested in deep class hierarchies,
adding unnecessary layers of complexity
for developers wishing
to extend those NEF components.
As a representative example,
the most often used class for making
NEF models, \texttt{NEFEnsemble},
extends from five ancestor classes
and implements seven interfaces.
Adding support for synaptic plasticity
in \texttt{NEFEnsemble}
involved adding or significantly modifying
25 Java classes \cite{TODO}.

@@

The JavaNengo scripting interface,
described in \cite{TODO},
provided a simpler interface
to the NEF-specific portions of the simulator.
This greatly enhanced productivity,
and made it possible
to create many models with very little code.
Modellers quickly started using
the scripting interface
as their primary tool,
but still needed to access the GUI
for interactive model inspection,
and still needed to modify the simulator
to add new functionality
(e.g., to add new neuron models).
This is because Jython makes it easy
to inspect and create Java objects,
but does not make it easy to
write Java code to
inspect and create Jython objects.
Modellers, therefore, had to be
proficient in Java whether
they used the scripting interface or not.
As there are few neural modellers
that are proficient with both Java and Python syntax,
the scripting interface still lacks
convenient ways to record the results of simulations,
and to define complex experimental setups.

Finally, perhaps the most significant limitation
of the JavaNengo tools is their performance.
While Java has good support for threads,
it does not have well-supported libraries
for doing mathematical operations on large matrices quickly.
Additionally, it is difficult to interface with
platform-specific tools such as
graphical processing units (GPUs),
which can perform the types of computations
that neural models require quickly and in parallel.

Addressing all of these issues
in the Java codebase would involve
significant enough changes
that we started from scratch instead.
Since the scripting interface
already used Python syntax,
and since Python is widely used
for neuroscientific data analysis,
we targeted the CPython interpreter
as the platform for the next generation of Nengo.

@@

PyNengo,
the subject of the remainder of this paper,
is a Python package for
defining and simulating
neural models using the NEF.
It targets the CPython interpreter,
meaning that it integrates seamlessly
with other scientific Python tools.
Arbitrary CPython programs
can use Nengo models,
opening up possibilities
for using neurally implemented algorithms
in applications and games.

PyNengo's object model is simple and minimal,
making it easy to
document, test, and modify.
It also includes a scripting interface to those objects
based on JavaNengo's scripting interface \cite{TODO},
but with several modifications
to allow for easier incorporation
of user-defined networks,
new neuron models, and learning rules.
Additionally, PyNengo's scripting interface
contains methods to record simulated data
and access that data after simulation.

Model creation and simulation are decoupled,
allowing for models to be exported
to other simulators,
such as those accessible with PyNN.
Simulators can also be written
to leverage specific platforms.
Currently, we have implemented
a platform-independent simulator
that uses NumPy for vectorized computations,
and a PyOpenCL-based
simulator that takes full advantage of GPUs
and multicore CPUs.

PyNengo does not currently
replace all of JavaNengo's functionality.
While it can simulate the vast majority
of models that JavaNengo can simulate,
some functionality is still in development.
Most notably, there is no graphical interface
to PyNengo, and therefore simulations
cannot be visualized while they are running.
However, we are confident that these tools
can be created, and will be much easier
to implement compared to JavaNengo's GUI.

PyNengo overcomes all of JavaNengo's limitations.
It can simulate most models
currently simulated by JavaNengo
using 11\% as many lines of code,
and depending on
only three external libraries instead of ten.
The OpenCL simulator can simulate
a large model TODO times
faster than JavaNengo's simulator.
This, along with a codebase
that is well tested and
easy to understand and extend,
provides a platform
that can simulate larger and more complex
models than Spaun,
and can therefore further
test the NEF as a theory of the brain.

@@

The Neural Engineering Framework (NEF; \cite{TODO})
proposes three principles
that enable the construction
of large-scale neural models.

\begin{enumerate}
  \item \textbf{Representation:} A population of neurons
    collectively represents information in the form of
    time-varying vectors of real numbers.
  \item \textbf{Transformation:} Functions on those vectors
    are computed by the connections between populations.
  \item \textbf{Dynamics:} The vectors represented
    by neural populations can be considered state variables
    in a dynamical system, and can be analyzed using control theory.
\end{enumerate}

@@

\paragraph{Representation}
Information is encoded by populations of neurons.
We consider that information
to be time-varying vectors of real numbers,
in order to reason with that information
using conventional mathematics.
By doing this, we can \textit{encode}
those vectors by injecting
specific amounts of current into
single neuron models based on
the vector being encoded.
This drives the neuron,
causing it to spike.
We can then estimate
the originally encoded vector
through a \textit{decoding} process.
This idea is similar to population coding
\cite{TODO}, but generalized
to vectors of arbitrary length.

Figure~\ref{fig:nef} depicts
the encoding (A) and decoding (B) process.
In the encoding process, the input signal drives
each neuron based on its \textit{tuning curve},
which describes how much
that neuron fires for a given input signal.
The tuning curve summarizes the gain
of a neuron (how quickly the activity rises),
the bias (the activity of a neuron given no signal),
and the encoding weight
(the part of the vector space
that causes the neuron to fire most strongly).
This results in a predictable pattern of firing
given the input signal.

In the decoding process,
the spike trains are first filtered
with an exponentially decaying filter.
Those filtered spike trains are summed together
with weights that are determined
by solving a least-squares minimization problem.
In the bottom panel of Figure~\ref{fig:nef}B,
the red line is the filtered spiking activity
weighted by decoding weights
that minimize the difference between
the actual encoded vector
and the decoded estimate.
The purple line is the filtered spiking activity
weighted by decoding weights
that minimize the different between
the cosine of the actual encoded vector
and the decoded estimate.
Note that these decoding weights are not determined
based on the input signal;
instead, the minimization occurs
on points that are randomly sampled
in the vector space
that the population represents.

\paragraph{Transformation}
Neurons communicate through
unidirectional connections called synapses.
When a neuron spikes,
it releases neurotransmitter across the synapse,
which causes some amount of current
to be imparted in the downstream neuron.
Many factors affect the
amplitude of the current imparted;
we summarize those factors
in a scalar connection weight
representing the strength
of the connection between two neurons.
In order to compute any function,
we can set the connection weights between
two populations to be the product of
the decoding weights for that function
for neurons in the first population
and the encoding weights
for the downstream neuron.

In Figure~\ref{fig:nef}C,
a sine wave is transformed twice.
Between populations A and B,
the signal is negated.
Between populations B and C,
the signal is squared.

Note that, in practice, we rarely use
the full connection weight matrix,
and instead store
the encoding and decoding weights
(i.e., the factors of the connection weight matrix).
This provides significant
space and time savings.

\paragraph{Dynamics}
While doing feedforward transformations
on vectors is sufficient to model
many neural systems,
many systems, notably memory systems,
require persistent activity through recurrent connections.
When recurrent connection are introduced,
the vectors that populations represent
can be thought of as state variables
in a dynamical system.
Neural systems can therefore
be built and analyzed using
the methods of control theory.

In Figure~\ref{fig:nef}D,
a population represents a two-dimensional vector,
and is recurrently connected
with negative feedback between dimensions,
resulting in a harmonic oscillator.

\paragraph{Nengo}
Large models can be built
by thinking of the principles of the NEF
as building blocks that can be put together
to describe neural systems.
The goal of Nengo is to allow
modellers to describe neural systems
in terms of what information is being represented,
how that information is transformed,
and what dynamics are required
to perform cognitively relevant
functions with biologically constrained
neural components.
Those descriptions are then
translated to a network
of interconnected neurons.
This situates Nengo
as a ``neural compiler''
that translates
a high-level functional model
to a low-level neural model
that implements the high-level function.

\section{PyNengo object model}

\subsection{Model}

A Nengo model is a collection
of Nengo objects that represent information
and connections that transform information.
Objects and connections can be probed
in order to collect data during simulation.

The \texttt{Model} object is primarily a container
for Nengo objects,
but also doubles as a scripting interface
that simplifies creating, connecting,
and probing objects.
Basic models
can be created solely through interacting
with the \texttt{Model} object
(see sections \ref{sec:comm-channel}--\ref{sec:lorenz}
for examples).
In advanced models,
objects are instantiated
and then added to the model
(see section \ref{sec:cconv} for an example).

\subsection{Ensemble}

An \texttt{Ensemble} is
a population of neurons
that represents information
in the form of a real-valued vector.
An ensemble must be given as arguments
its dimensionality
(i.e., the length of the vector it represents),
and an object that describes
the population of neurons.
For example, \texttt{nengo.LIF(50, tau\_ref=0.002)}
describes a population
of 50 leaky integrate-and-fire neurons
with a refractory time of 0.002 seconds.
Neurons are defined symbolically
so that each simulator can compute
this population's nonlinear function
however it sees fit.
The neuron model used by the population
is easily changed, as long as the simulator
is aware of the neuron model description object.

Other attributes of the \texttt{Ensemble},
such as its encoding weights,
the maximum firing rate of its neurons,
and so on, can also be specified
either as keyword arguments
to the \texttt{Ensemble} constructor,
or by setting an attribute on the instantiated object.
While an ensemble makes a hypothesis
about the information being represented by neurons,
these additional attributes
also allow modellers to ensure that
the ensemble stays
within neurobiological constraints.
If these attributes are not set,
then they will be randomly selected
from distributions that describe
neocortical pyramidal cells.

\subsection{Node}

A \texttt{Node} tracks any information
that is not represented by an ensemble.
In the most common case,
nodes provide input signals
that drive ensembles of sensory neurons.
In a more general sense,
nodes represent the experimental environment
in which a neural model exists.

@@

Nodes allow many of the special-purpose
routines that are necessary in real-world
models to be integrated
as core components of a Nengo model.
This makes Nengo models more explicit,
and enables simulators
to control and optimize node execution.

\subsection{Connection}

@@

The most important type of connection
in Nengo is the \texttt{DecodedConnection}.
This connection implements
the NEF's transformation principle.
In other words, the \texttt{DecodedConnection}
allows ensembles to project
the information they're representing---or
a transformation of that information---to
another ensemble or node.
This functionality is what enables Nengo models
to appear conceptual,
even though the underlying implementation
can translate that connection
to connection weights.

The ability to connect neurons together
with synaptic connection weights is still possible
with the \texttt{NonlinearityConnection}.
Both of these connection types
can be temporally filtered,
and the weights involved in the connection
(decoding weights for \texttt{DecodedConnection},
synaptic connection weights for \texttt{NonlinearityConnection})
can be modified over time with learning rules.

In many cases, the type of the connection
is not explicitly specified by the modeller.
Instead, it is inferred by the arguments
to \texttt{Model}'s \texttt{connect} function.
However, that function is just a simple wrapper
of each connection object's constructor,
so complex models can use
the appropriate connection,
or even implement their own.

\subsection{Network}

A network is a collection of ensembles and nodes
connected together in a particular way.
It is primarily a way of grouping together
a common set of connected objects,
and makes explicit the purpose
of the network and how it interacts
with the rest of the model
(see section \ref{sec:ccon} for a usage example).

Networks are designed to be
subclasses by users,
in order to group together
parts of a model.
In most cases,
the code that creates and connects
several objects in the model can be
moved into a \texttt{Network}
subclass' \texttt{make}
function with only minor changes.

Networks can vary dramatically in size,
and can be composed of other networks.
The \texttt{Integrator} network, for example,
is composed of only one recurrently connected ensemble.
By encapsulating that logic in a network,
the purpose of that ensemble is made explicit.
On the other end of the spectrum,
the \texttt{BasalGanglia} network,
is composed of five groups of ensembles
connected with several specific functions
that together implement a ``winner take all'' circuit.
Encapsulating that logic
makes a complicated section of code
easy to include in many different models.

\subsection{Probe}

A probe is used to denotes a signal
that will be recorded
throughout the simulation.
Nengo objects and connections
each have several signals that can be probed.
An ensemble's decoded value
and its neurons' spikes
can be probed, among others.

Like nodes, a probe could be implemented
outside of the model.
However, doing so requires detailed knowledge
of the simulator,
and can incur significant overhead
if not implemented carefully.
For these reasons, we have made probes
a core component of a Nengo model,
and are therefore explicit
and optimizable.
Further, since probes are described
at a symbolic level,
the underlying implementation
can output probed data in many different formats.
Currently, simulators store probed data
directly in memory,
but adding the ability to store data
in files or to stream probed data
directly to sockets will be added.

\section{PyNengo simulators} \label{sec:simulators}

\begin{figure}
\begin{center}
  \includegraphics[width=0.6\textwidth]{sim}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:sim} Figure \arabic{figure}.}{
   The first two timesteps of a simple simulation.}
\end{figure}

Decoupling model creation and simulation
enables PyNengo simulators
to allocate memory and schedule computations
in the most efficient ways possible.
Simulators are given a \texttt{Model}
as an argument;
this \texttt{Model} is a symbolic description,
which a simulator can assume will not change
over the course of a simulation.
The simulator, however,
can modify the model as it sees fit;
in most cases, this means that the simulator
will fill in many of the details
that the symbolic description
has allowed to be generated randomly.

We have implemented
a reference simulator using NumPy
for vectorized computations,
and an OpenCL simulator
using PyOpenCL to parallelize
computations on GPUs and multicore CPUs.
In the remainder of this section,
we will describe
the reference simulator implementation;
the OpenCL simulator shares many
of the reference simulator's architectural choices,
but is heavily optimized in ways
that cannot be adequately described
with the space available.
Additionally, it should be noted that
any simulator that takes a \texttt{Model}
as input and exposes a \texttt{step}
function can be used in PyNengo;
the reference simulator
is provided as an example
for simulator designers,
not as a specification.

@@

Before the first timestep, the reference simulator
\begin{enumerate}
  \item translates high-level objects to
    a set of signals and operators,
  \item allocates NumPy arrays for each signal, and
  \item sorts operators based on a dependency graph.
\end{enumerate}
On each timestep, the reference simulator
\begin{enumerate}
  \item computes each operator in order, and
  \item copies probed signals to a separate buffer.
\end{enumerate}
Figure~\ref{fig:sim} depicts
the state of the reference simulator
after two timesteps of a simple model.

\paragraph{Signals}

A \texttt{Signal} represents any number that
will be used by the simulator.
Each high level Nengo object contains
several signals;
for example, an ensemble contains signals
that represent the high-level input
signal that will be encoded
to input currents,
and the encoding weights.
It also contains a neural population,
which contains signals that represent
input currents, bias currents,
membrane voltages, and refractory times for each cell.
